---
title: "K-medoids Distance-Based clustering"
author: "Weksi Budiaji"
date: "`r Sys.Date()`"
output: html_vignette
vignette: >
  %\VignetteIndexEntry{K-medoids Distance-Based clustering}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## 1. Introduction

The **kmed** package was designed to analyse k-medoids based clustering; the features include:

- distance computation (numerical, binary, categorical, and mixed variables)
- k-medoids algorithms
- bootstrap evaluation and visualization


## 2. Distance Computation

### A. Numerical variables

There are four distance options, i.e. manhattan weighted by rank (`mrw`), squared euclidean weighted by rank (`ser`) and variance (`sev`), and squared euclidean unweighted (`se`). The `distNumeric` in the `method` provides its desired distance method.
```{r}
library(kmed)
num <- as.matrix(iris[,1:4])
mrwdist <- distNumeric(num, num, method = "mrw")
mrwdist[1:6,1:6]
```

### B. Binary and Categorical variables

Two options of distances are available for binary or categorical variables, namely simple matching (`matching`) and coocurrence (`coocurrence`) distance.
```{r}
set.seed(1)
a <- matrix(sample(1:2, 7*3, replace = TRUE), 7, 3)
matching(a, a)
coocurance(a)
```

### C. Mixed variables

There are five distances for mixed variables data, i.e gower (`gower`), wishart (`wishart`), podani (`podani`), huang (`huang`) and harikumar-pv (`harikumar`).
```{r}
a1 <- matrix(sample(1:3, 7*3, replace = TRUE), 7, 3)
mixdata <- cbind(iris[1:7,1:3], a, a1)
colnames(mixdata) <- c(paste(c("num"), 1:3, sep = ""), paste(c("bin"), 1:3, sep = ""), paste(c("cat"), 1:3, sep = ""))
distmix(mixdata, method = "gower", idnum = 1:3, idbin = 4:6, idcat = 7:9)
distmix(mixdata, method = "wishart", idnum = 1:3, idbin = 4:6, idcat = 7:9)
```

## 3. K-medoids algorithms

There are some k-medoids algorithms, partitioning around medoids, for example, is available in [**cluster**](https://cran.r-project.org/package=cluster) package. For this moment, in **kmed** package, the available algorithm is park and jun.
```{r}
result <- fastkmed(mrwdist, ncluster = 3, iterate = 50)
table(result$cluster, iris[,5])
```

## 4. Bootstrap evaluation

### A. Bootstrap replicate matrix

To evaluate the clustering algorithm, a bootstrap evaluation can be run by make a function. This function input arguments must be only a distance matrix and a number of cluster and the output is only membership.
```{r}
k <- 3
parkboot <- function(x, nclust) {
  res <- fastkmed(x, nclust, iterate = 50)
  return(res$cluster)
}
irisboot <- clustboot(mrwdist, nclust=k, parkboot, nboot=50)
irisboot[1:5,c(1:5,46:50)]
```
We can change the algorithm, for example `kmeans` from **stats** package.
```{r}
kmboot <- function(x, nclust) {
  res <- kmeans(x, nclust)
  return(res$cluster)
}
kmboot <- clustboot(num, nclust=k, kmboot, nboot=50, diss = FALSE)
kmboot[1:5,c(1:5,46:50)]
```


### B. Consensus matrix

A consensus matrix  (n x n) can be produced from a bootstrap replicate matrix. The `reorder` input is a function to reorder the objects in the consensus matrix. This function input arguments must be only a distance matrix and a number of cluster and the output is only membership. This matrix can be visualized using heatmap directly.
```{r}
wardorder <- function(x, nclust) {
  res <- hclust(x, method = "ward.D2")
  member <- cutree(res, nclust)
  return(member)
}
consensusiris <- consensusmatrix(irisboot, nclust = k, wardorder)
consensusiris[c(1:5,51:55,101:105),c(1:5,51:55,101:105)]
```

### C. Visualization (Heatmap)

To produce a heatmap of consensus matrix `clustheatmap` can be applied in the consensus matrix. The consensus matrix heatmap of Iris data by Park and Jun is produced.
```{r}
clustheatmap(consensusiris, "Iris Data via Park and Jun")
```

We can also create a heatmap of the kmeans algorithm for the iris data.
```{r}
consensusiris2 <- consensusmatrix(kmboot, nclust = k, wardorder)
clustheatmap(consensusiris2, "Iris Data via kmeans")
```

